---
title: "Data overviews"
author: "Andrew Irwin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)
library(lterdatasampler) # remotes::install_github("lter/lterdatasampler")
```

When you first get a new dataset into R you should immediately have a few questions: 

* What is in this dataset?
* Was the data read correctly?
* Are there any easily spotted errors?

The following functions give quick summaries of the data that may help you with these questions.

* head, str, pillar::glimpse, vtree, naniar

```{r}
head(penguins)
```

```{r}
str(penguins)
pillar::glimpse(penguins)
```




```{r}
vtree::vtree(penguins, "species island sex", sameline = TRUE)
vtree::vtree(penguins, "year island", sameline = TRUE)
```

## Where are those missing values?

The package `naniar` has many functions to identify and summarize missing data in a data frame. Here are just a few examples.

```{r}
naniar::miss_var_summary(penguins)
naniar::miss_case_table(penguins)
naniar::gg_miss_upset(penguins)
```


