---
title: "Date calculations"
author: "Andrew Irwin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
```

Here are the tasks I do most frequently with dates:

* Convert text into a date
* Extract parts of a date (the year, month, day, week, day of year, convert date to a decimal year)
* Find the length of time between two dates

I use the `lubridate` package for all these tasks. Related proejcts `date` and `chron` may be helpful too.

It's helpful to know that dates, times, and combinations of dates and times are different in R. Furthermore, there are different ways to represent dates.

## Common tasks

### Converting text to a date

If you have a date in a column of data, when you read it into R with `read_csv`, the function will attempt to recognize the text as a date and convert it for you. Since some date formats are ambiguous, this will not always be possible. (Create an example file.) The family of functions `ymd`, `dmy`, `mdy` recognize text dates formatted in that order to dates. If the date is illegal (e.g., April 31 or February 29 in a non-leap year) the result will be NA.

```{r}
ymd(c("2022-06-12", "2022 June 12", "22/6/12"))
dmy(c("12/6/22", "12th of June, 2022.", "12-06-2022"))
mdy(c("6/12/22", "June 12, 2022", "06-12-2022"))
```

If you write months in a language that is not part of your locale (see `Sys.getlocale()` and `readr::locale()`) then the results may surprise you. Can you tell what happened here?

```{r}
mdy(c("Juin 12, 2022",  "Junio 12, 2022"))
```

If you have a lot of dates in a strange format that these functions can't interpret, you can use `parse_date_time` to describe what you want done.

### Extracting parts of a date

```{r}
n = today()
year(n)
month(n)
day(n)
week(n)
quarter(n)
yday(n)
leap_year(n)
decimal_date(n)
```

### Creating dates

```{r}
now()
today()
make_date(2022, 6, 12)
date_decimal(2022.4447, tz = "America/Halifax")
as_date(135)  # number of days since Jan 1, 1970
```
