---
title: "Tidy Tuesday"
date: 2022-06-28
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The way we organize data affects how we think about the data and the tools we use to analyze and visualize the data. Many people advocate a particularly simple organizational style known as "tidy data": each column is a different variable, with the same units in all rows. Each row is an observation and the entries for all columns in that row refer to the same observation.

The advantages of [tidy data](https://tidyr.tidyverse.org/) are discussed in the descriptions of the "tidyverse", especially "ggplot" and "dplyr" tools. For more see the [R for Data Science](https://r4ds.had.co.nz/) book.

A team of interested volunteers is developing a curated set of tidy data from a wide range of topics, with new data sets released weekly. For alliterative fun this is called the [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday) project. The idea is that you can pick the weekly data, or any data you like, and start exploring it as a way to motivate teaching yourself (or learning in a group) to analyze and visualize data.

# Tidy Tuesday

Tidy Tuesday are not always tidy! You might want to apply various R for Data Science techniques to wrangle the data into a true tidy format.
The goal of TidyTuesday is to apply your R skills, get feedback, explore other’s work.

To get started, use the `tidyverse` and `tidytuesdayR` packages.

```{r}
library(tidyverse, quietly = TRUE)
library(tidytuesdayR)
```

# Load the weekly Data

You can get data from any week using `tt_load`.

```{r Load}
tt <- tt_load("2019-10-15")
# tt <- tt_load("2022-06-21")
```

You can also create a new R markdown file for tidy data using the `use_tidytemplate` function in the console. Try this function call:

```{r eval=FALSE}
use_tidytemplate()
```

# Readme

Take a look at the readme for the weekly data to get insight on the dataset.
This includes a data dictionary, source, and a link to an article on the data. (The information does not show up in the knitted output, but if you view the data in Rstudio you'll see information about the data in the 'viewer' window.)

```{r Readme, eval = interactive()}
tt 
```


# Glimpse Data

You can glimpse the data using the `glimpse` function. (Here I use `map` to glimpse each dataset in tt.)

```{r Glimpse}
tt %>% 
  map(glimpse)
```

# Wrangle

Explore the data and process it into a nice format for plotting! Access each dataset by name by using a dollarsign after the `tt` object and then the name of the data set. You can see the names of the datasets in `tt` using `names`:

```{r Wrangle}
names(tt)
tt$big_epa_cars
```


# Visualize

Using your processed dataset, create your unique visualization.

```{r Visualize}
tt$big_epa_cars |>
  ggplot(aes(fuelCost08, 
             fct_reorder(drive, fuelCost08))) +
  geom_boxplot()
```

# Save Image

Save your image for sharing. You can save your most recent plot using `ggsave`:

```{r}
ggsave(filename = "My TidyTuesday Plot.png")
```
